MISUSES AND INACCURACIES OF PAGEL'S LAMBDA ML ESTIMATION METHODS

### 2020-04-20 Manuscript/Manuscript-1-DCA.Rmd
vsn 1

  In this study, we evaluate the precision of Pagel's $\lambda$ in estimating known levels of phylogenetic signal in phenotypic data. First we use computer simulations across differing numbers of species, differently shaped phylogenies, and differing input levels of phylogenetic signal, to explore the degree to which $\lambda$ correctly identifies known levels of phylogenetic signal, and under what circumstances. For comparison, we also evaluate estimates of $Kappa$ and its empirically-derived effect size ($Z_K$) for this purpose. Additionally, we use simulations to determine how the inclusion of $\lambda$ in phylogenetic regression and ANOVA (i.e., PGLS) affects parameter estimation, and whether levels of phylogenetic signal estimated in a PGLS framework are accurate. We then survey the recent macroevolutionary literature for published papers containing estimates of $\lambda$ from empirical datasets, and compare these empirical estimates to patterns gleaned from our computer simulations. In general we find that while PGLS parameters (e.g., $\beta$) are accurately estimated with the inclusion of phylogenetic signal, estimates of $\lambda$ are not. Further, we find that estimates of $\lambda$ vary widely for a given input value of phylogenetic signal, and that this variation is not constant across the range of input signal, with decreased precision when phylogenetic signal is of intermediate strength. By contrast, variation across effect sizes ($Z_K$) obtained from $Kappa$ is far more consistent across the range of input values, and that the precision of estimating $Z_K$ is preferable to that observed using $\lambda$. We conclude that estimates of phylogenetic signal using Pagel's $\lambda$ are often innacurate, and thus interpreting strength of phylogenetic signal in phenotypic datasets based on this measure should be treated with caution. By contrast, effect sizes obtained from $Kappa$ hold promise for characterizing phylogenetic signal, and comparing the strength of phylogenetic signal across datasets.

@@

We conducted a series of computer simulations to evaluate the ability of Pagel's [-@Pagel1999] $\lambda$ to estimate known levels of phylogenetic signal. Our primary simulations were based on pure-birth phylogenies; however, we also evaluated patterns on both balanced and pectinate trees to determine whether tree shape affected our findings (see Supporting Information). Our first set of simulations evaluated the extent to which values of $\lambda$ estimated from the data corresponded with actual input levels of phylogenetic signal. For these simulations we generated 50 pure-birth phylogenies at each of six different tree sizes, ranging from 32 to 1024 taxa ($n=2^5 - 2^{10}$). Next, we rescaled the simulated phylogenies by multiplying the internal branches by $\lambda$. We used a set of values that encompassed the entire range of $\lambda$: ($\lambda_{in} = 0.0 \to 1.0$; in 21 intervals of 0.05 units), resulting in 1050 scaled phylogenies at each level of species richness ($n$). Continuous traits were then simulated on each phylogeny under a Brownian motion model of evolution to obtain datasets with known and differing levels of phylogenetic signal. These varied from datasets with no phylogenetic signal (when $\lambda_{in} =0$) to datsets with phylogenetic signal corresponding to what was expected under Brownian motion (when $\lambda_{in} =1$). \hfill\break

Using the simulated phylogenies we estimated the phylogenetic signal in each dataset using Pagel's $\lambda$. Next, for the 1050 datasets at each level of species richness ($n$), we assessed the relationship between $\lambda_{in}$ and $\lambda_{est}$ using regression. Additionally, the precision of $\lambda_{est}$ was approximated by observing the variation of $\lambda_{est}$ obtained at each input level of phylogenetic signal ($\lambda_{in}$). Finally, for comparison we estimated the degree of phylogenetic signal in each dataset using $Kappa$ [@Blomberg_et_al2003; @Adams2014a], and from this calculated its effect size [$Z_K$: sensu @AdamsCollyer2016; @AdamsCollyer2019b] for each dataset. Here, the effect size was estimated as: $Z_K=\frac{K_{obs}-\mu_K}{\sigma_K}$, where $K_{obs}$ was the observed $Kappa$, and $\mu_K$ and $\sigma_K$ were the mean and standard deviation of the empirical sampling distribution of values obtained from the permutation distribution [for discussion see: @AdamsCollyer2019]. Because $Z_K$ differs in scale from $\lambda$, we used a linear normalization to standardize $Z_K$ to a uniform distribution ($0\rightarrow1$), and variation in the set of normalized $Z_K$ values at each input level of phylogenetic signal ($\lambda_{in}$) were then estimated for comparison. \hfill\break

Our second set of simulations evaluated the extent to which values of $\lambda$ estimated in PGLS regression and ANOVA (i.e., $Y\sim{X}$) corresponded with actual input levels of phylogenetic signal in the response variable. As before we generated 50 pure-birth phylogenies at each of six levels of species richness ($n=2^5 - 2^{10}$), and rescaled each with a set of input $\lambda$ values ($\lambda_{in} = 0.0 - 1.0$; in 21 intervals of 0.05 units). Next, an independent variable $X$ was simulated on each phylogeny under a Brownian motion model of evolution (for PGLS regression). For phylogenetic ANOVA, random groups ($X$) were obtained by simulating a discrete (binary) character on each phylogeny. Next, the dependent variable was simulated in such a manner as to contain a known relationship with $X$ plus random error containing phylogenetic signal. This was accomplished as: $Y=\beta{X}+\epsilon$. Here, the association between $Y$ and $X$  was modeled using a range of values: $\beta=(0.0,0.25, 0.5, 0.75,1.0)$, and the residual error was modeled to contain phylogenetic signal simulated under a Brownian motion model of evolution: $\epsilon=\mathcal{N}(\mu=0,\sigma=\mathbf{C})$: [see @Revell2010 for a similar simulation design]. For each dataset, the fit of the phylogenetic regression was estimated using maximum likelihood, and parameter estimates ($\beta_{est}$ and $\lambda_{est}$) were obtained and evaluated as above. All analyses were performed in R v3.6.0 [@RCT] using the packages `geiger` [@Harmon2008], `caper` [@Orme2013], `phytools` [@Revell2012], and `geomorph` [@AdamsOtarola2013; @AdamsGeomorph]. R-scripts are found in the Supplemental Material.

@@

lambda: (Fig 1) not good predictor of input physignal (regression), particularly at low sample sizes. Also, similar to early results of Boettiger (but greatly expanded) range and variation of lambda very large at low N, and generally decreases with larger N. However, the precision was not uniform across the range of lambda-in. Variation in lambda-est was greatest at intermediate levels of physig, and descreased at lower and higher values. Thus, the overall dispersion across the range of input values resembled an asymmetric lens (**NAME?**), with less variation at lambda-in close to 1, greatest variation in middle, and less (but substantial) variation near lambda-in of zero.  Notably,  even at very large species richness (n=1024) the range of lam-est was quite large: e.g., l-in 0.5 has range of 45% of all possible values. Thus, lambda-est is not very precise in terms of estimating the true l-in parameter, meaning that using lambda to biologically interpret the 'strength' of phylogenetic signal in a dataset, or compare between datasets, is fraught with difficulty. \hfill\break

@@

By contrast,  (FIg 2) Kappa. Much smaller range and variation as compared with lambda, but with larger l-in, variation in kappa does increase. However, this is not the case with Z-K.  Z-K increases predictably (linearly, progressively...) with L-in, and importantly variation in Z-K remains nearly constant across the range of input values. this means that Z-K is equally precise in estimating the strength of phylogenetic signal, and may be more adept at facilitating useful statements on teh relative strength of phylogenetic signal within and across datasets. \hfill\break

@@

2: expand on Lambda..
lambda innacurate, not precise, level of precision varies with input physig (worse in mid-range). Patterns worse with PGLS, though beta still estimated properly. Conclusion, lambda not overly useful.

@@

\textbf{Figure 2}. Variation in estimates of phylogenetic signal across input levels of phylogenetic signal. **A**: Estimates of Pagel's $\lambda$ for data simulated on phylogenies with 128 taxa ($n=128$), **B** Estimates of $Z_K$ for data simulated on phylogenies with 128 taxa ($n=128$), **C** Variance in the variation of $\lambda_{est}$ across input levels of phylogenetic signal, estimated on phylogenies containing differing numbers of species. **D** Variance in the variation of $Z_K$ across input levels of phylogenetic signal, estimated on phylogenies containing differing numbers of species.  \hfill\break

@@

\singlespacing \textbf{Figure 2}.  Variation in estimates of phylogenetic signal across input levels of phylogenetic signal. **A**: Estimates of Pagel's $\lambda$ for data simulated on phylogenies with 128 taxa ($n=128$), **B** Estimates of $Z_K$ for data simulated on phylogenies with 128 taxa ($n=128$), **C** Variance in the variation of $\lambda_{est}$ across input levels of phylogenetic signal, estimated on phylogenies containing differing numbers of species. **D** Variance in the variation of $Z_K$ across input levels of phylogenetic signal, estimated on phylogenies containing differing numbers of species.

>>>>

In this study, we evaluate the precision of Pagel's $\lambda$ in estimating known levels of phylogenetic signal in phenotypic data. First we use computer simulations across differing numbers of species, differently shaped phylogenies, and differing input levels of phylogenetic signal, to explore the degree to which $\lambda$ correctly identifies known levels of phylogenetic signal, and under what circumstances. For comparison, we also evaluate estimates of $Kappa$ and its empirically-derived effect size ($Z_K$) for this purpose. Additionally, we use simulations to determine how the inclusion of $\lambda$ in phylogenetic regression and ANOVA (i.e., PGLS) affects parameter estimation, and whether levels of phylogenetic signal estimated in a PGLS framework are accurate. We then survey the recent macroevolutionary literature for published papers containing estimates of $\lambda$ from empirical datasets, and compare these empirical estimates to patterns gleaned from our computer simulations. In general we find that while PGLS parameters (e.g., $\beta$) are accurately estimated with the inclusion of phylogenetic signal, estimates of $\lambda$ are not. Further, we find that estimates of $\lambda$ vary widely for a given input value of phylogenetic signal, and that this variation is not constant across the range of input signal, with decreased precision when phylogenetic signal is of intermediate strength. By contrast, variation across effect sizes ($Z_K$) obtained from $Kappa$ is far more consistent across the range of input values, and the precision of estimating $Z_K$ is preferable to that observed using $\lambda$. We conclude that estimates of phylogenetic signal using Pagel's $\lambda$ are often innacurate, and thus interpreting strength of phylogenetic signal in phenotypic datasets based on this measure should be treated with caution. By contrast, effect sizes obtained from $Kappa$ hold promise for characterizing phylogenetic signal, and comparing the strength of phylogenetic signal across datasets.

@@

We conducted a series of computer simulations to evaluate the ability of Pagel's $\lambda$ to estimate known levels of phylogenetic signal. Our primary simulations were based on pure-birth phylogenies; however, we also evaluated patterns on both balanced and pectinate trees to determine whether tree shape affected our findings (see Supporting Information). Our first set of simulations evaluated the extent to which values of $\lambda$ estimated from the data corresponded with actual input levels of phylogenetic signal. For these simulations we generated 50 pure-birth phylogenies at each of six different tree sizes, ranging from 32 to 1024 taxa ($n=2^5 - 2^{10}$). Next, we rescaled the simulated phylogenies by multiplying the internal branches by $\lambda$. We used a set of values that encompassed the entire range of $\lambda$: ($\lambda_{in} = 0.0 \to 1.0$; in 21 intervals of 0.05 units), resulting in 1050 scaled phylogenies at each level of species richness ($n$). Continuous traits were then simulated on each phylogeny under a Brownian motion model of evolution to obtain datasets with known and differing levels of phylogenetic signal. These varied from datasets with no phylogenetic signal (when $\lambda_{in} =0$) to datsets with phylogenetic signal corresponding to what was expected under Brownian motion (when $\lambda_{in} =1$). \hfill\break

Using the simulated phylogenies we estimated the phylogenetic signal in each dataset using Pagel's $\lambda$. Next, for the 1050 datasets at each level of species richness ($n$), we assessed the relationship between $\lambda_{in}$ and $\lambda_{est}$ using regression. Additionally, the precision of $\lambda$ was approximated by observing the variation of $\lambda_{est}$ obtained at each input level of phylogenetic signal ($\lambda_{in}$). Finally, for comparison we estimated the degree of phylogenetic signal in each dataset using $Kappa$ [@Blomberg_et_al2003; @Adams2014a], and from this calculated its effect size [$Z_K$: sensu @AdamsCollyer2016; @AdamsCollyer2019b] for each dataset. Here, the effect size was estimated as: $Z_K=\frac{K_{obs}-\mu_K}{\sigma_K}$, where $K_{obs}$ was the observed $Kappa$, and $\mu_K$ and $\sigma_K$ were the mean and standard deviation of the empirical sampling distribution of values obtained from the permutation distribution [for discussion see: @AdamsCollyer2019]. Because $Z_K$ differs in scale from $\lambda$, we used a linear normalization to standardize $Z_K$ to a uniform distribution ($0\rightarrow1$), and variation in the set of normalized $Z_K$ values at each input level of phylogenetic signal ($\lambda_{in}$) were then estimated for comparison. \hfill\break

Our second set of simulations evaluated the extent to which values of $\lambda$ estimated in PGLS regression and ANOVA (i.e., $Y\sim{X}$) corresponded with actual input levels of phylogenetic signal in the response variable. As before we generated 50 pure-birth phylogenies at each of six levels of species richness ($n=2^5 - 2^{10}$), and rescaled each with a set of input $\lambda$ values ($\lambda_{in} = 0.0 - 1.0$; in 21 intervals of 0.05 units). Next, an independent variable $X$ was simulated on each phylogeny under a Brownian motion model of evolution (for PGLS regression). For phylogenetic ANOVA, random groups ($X$) were obtained by simulating a discrete (binary) character on each phylogeny. Next, the dependent variable was simulated in such a manner as to contain a known relationship with $X$ plus random error containing phylogenetic signal. This was accomplished as: $Y=\beta{X}+\epsilon$. Here, the association between $Y$ and $X$  was modeled using a range of values: $\beta=(0.0,0.25, 0.5, 0.75,1.0)$, and the residual error was modeled to contain phylogenetic signal simulated under a Brownian motion model of evolution: $\epsilon=\mathcal{N}(\mu=0,\sigma=\mathbf{C})$: [see @Revell2010 for a similar simulation design]. For each dataset, the fit of the phylogenetic regression was estimated using maximum likelihood, and parameter estimates ($\beta_{est}$ and $\lambda_{est}$) were obtained and evaluated as above. All analyses were performed in R v3.6.0 [@RCT] using the packages `geiger` [@Harmon2008], `caper` [@Orme2013], `phytools` [@Revell2012], and `geomorph` [@AdamsOtarola2013; @AdamsGeomorph]. R-scripts are found in the Supporting Information.

@@

Our primary series of simulations revealed several patterns. First, the relationship between $\lambda_{est}$ and input levels of phylogenetic signal was $\sim\beta=1.0$ at each level of species richness ($n=32, 64, 128, 256, 512, 1024$). This implied that the mean $\lambda_{est}$ across datasets for a given set of simulation conditions reflected input levels of phylogenetic signal. However, as shown in Figure 1, the precision of those estimates varied widely, and in several interesting ways. Predictably, precision was worst at low levels of species richness, where in many cases the set of $\lambda_{est}$ spanned nearly the entire range of possible values (e.g., $n=32$; $\lambda_{est}=0.5$: range = $0.0\to0.985$). Second, as species richness increased, variation across estimates of $\lambda$ decreased (Figure 1). This confirmed the pattern identified by Boettiger et al. [-@Boettiger_et_al2012] on  a small ($n=13$) and large ($n=281$) phylogeny, and demonstrated that this trend was general across a wider range of species richness. Importantly however, our broader set of simulations revealed that the precision of $\lambda_{est}$ was not uniform across all levels of phylogenetic signal, and that there was a particular pattern to the imprecision of estimating this phylogenetic parameter. Specifically, we found that variation in $\lambda_{est}$ was highest at intermediate levels of phylogenetic signal, and decreased at both lower and higher levels of input signal (Figure 1). This implied that the precision in estimating $\lambda$ was worst at intermediate levels of phylogenetic signal; a pattern that remained consistent across all levels of species richness evaluated here (Figure 1). Notably, even at very large levels of species richness, the range of $\lambda_{est}$ still encompassed a substantial portion of possible values (e.g., $n=1024$; $\lambda_{est}=0.5$: range = $0.31\to 0.68$). From this we conclude that $\lambda_{est}$ is not an accurate estimator of the strength of phylogenetic signal in phenotypic datasets, and that biological interpretations of the level of phylogenetic signal based on $\lambda$ should be treated with caution. \hfill\break

@@

**STOPPED HERE**By contrast,  (FIg 2) Kappa. Much smaller range and variation as compared with lambda, but with larger l-in, variation in kappa does increase. However, this is not the case with Z-K.  Z-K increases predictably (linearly, progressively...) with L-in, and importantly variation in Z-K remains nearly constant across the range of input values. this means that Z-K is equally precise in estimating the strength of phylogenetic signal, and may be more adept at facilitating useful statements on teh relative strength of phylogenetic signal within and across datasets. \hfill\break

@@

2: expand on Lambda..
lambda innacurate, not precise, level of precision varies with input physig (worse in mid-range). NEW RESULT.  We are first to show this.  NOTE: pattern is obvious with reflection. Since it is a 'bounded' parameter estimation should be best at the extremes... (state this?).. hmm.

Patterns worse with PGLS, though beta still estimated properly. Conclusion, lambda not overly useful.

@@

\textbf{Figure 2}.  Variation in estimates of phylogenetic signal across input levels of phylogenetic signal. (A) Estimates of Pagel's $\lambda$ for data simulated on phylogenies with 128 taxa ($n=128$), (B) Estimates of $Z_K$ for data simulated on phylogenies with 128 taxa ($n=128$), (C) Variance in the variation of $\lambda_{est}$ across input levels of phylogenetic signal, estimated on phylogenies containing differing numbers of species. (D) Variance in the variation of $Z_K$ across input levels of phylogenetic signal, estimated on phylogenies containing differing numbers of species.    \hfill\break

@@

\singlespacing \textbf{Figure 2}.  Variation in estimates of phylogenetic signal across input levels of phylogenetic signal. (A) Estimates of Pagel's $\lambda$ for data simulated on phylogenies with 128 taxa ($n=128$), (B) Estimates of $Z_K$ for data simulated on phylogenies with 128 taxa ($n=128$), (C) Variance in the variation of $\lambda_{est}$ across input levels of phylogenetic signal, estimated on phylogenies containing differing numbers of species. (D) Variance in the variation of $Z_K$ across input levels of phylogenetic signal, estimated on phylogenies containing differing numbers of species.
